### 2020-03-04 doc/icfp20/conclusion.tex
final tweaks & swap figures

We believe that functional programming, and in particular functional
programming with dependent types, is a promising way to do software
engineering going forward. The languages and tools are perhaps not yet
as finely polished and rich as those typically used in production
software, but they are fundamentally far more powerful, and thus once
the infrastructure (and training of programmers) catches up should
become the languages of choice.

Music is an especially valuable domain in which to explore these
ideas. We have found it to be a microcosm of issues that arise
everywhere in software engineering: modularity, composability,
correctness, and data issues such as handling equivalent formulations of
the same data as well slightly different formulations. If we can
demonstrate that functional programming with dependent types works
well in the domain of music, the same techniques can be used in
software in general.

@@

For now we have explicitly converted between equivalent representations
and explicitly lifted functions, to get a feel for the amount of work
required.  However in the world of dependent types there are known
research techniques for handling these cases, namely transporting
across equivalences as in Homotopy Type Theory~\citep{hottbook} and
ornaments~\citep{dagand-ornaments}. Cubical
Agda~\citep{vezzosi-cubical}, although still early in development,
should be an excellent tool for exploring the extent to which these
research ideas can be applied in a practical context.

>>>>

% We believe that functional programming, and in particular functional
% programming with dependent types, is a promising way to do software
% engineering going forward. The languages and tools are perhaps not yet
% as finely polished and rich as those typically used in production
% software, but they are fundamentally far more powerful, and thus once
% the infrastructure (and training of programmers) catches up should
% become the languages of choice.

In this paper we have reported our experience in using Agda for
formalizing music theory.
We have found music to be a microcosm of issues that arise
everywhere in software engineering: modularity, composability,
correctness, and data issues such as handling equivalent or slightly
different formulations of the same data. If we can
demonstrate that functional programming with dependent types works
well in the domain of music, the same techniques can be used in
software in general.

@@

For now we have explicitly converted between equivalent representations
and lifted functions, to get a feel for the amount of work
required.  However in the world of dependent types there are known
research techniques for handling these cases, namely transporting
across equivalences as in Homotopy Type Theory~\citep{hottbook} and
ornaments~\citep{dagand-ornaments}. Cubical
Agda~\citep{vezzosi-cubical}, although still early in development,
should be an excellent tool for exploring the extent to which these
research ideas can be applied in a practical context.

### 2020-03-04 doc/icfp20/counterpoint.tex
add scale degree & inversion to glossary
final tweaks & swap figures

In this section, we present an implementation of species counterpoint,
based on the formulation given by \citet{fux-cp}.
The idea is to represent ``good'' counterpoint  as a dependent record,
whose fields encode the musical content as well as proofs that the
counterpoint follows certain rules.
For space reason, we only describe two variants of species counterpoint;
other variants can be formalized in an analogous way.

@@

\begin{alltt}  
first : PitchInterval -- Pitch \(\times\) Interval
first = (c 5 , per8)

@@

The middle bars of the music should maintain consonance of intervals
and independence of melodic lines.
As we saw in Figure~\ref{fig:interval}, consonant intervals include
the unison, 3rd, 5th, 6th, and 8th, and among these, the unison is
clearly an obstacle to distinguishing between the two lines of music.
Therefore, the middle bars must consist only of the latter four intervals.
We encode this rule as the \texttt{checkIntervals} function, which
returns a list of errors corresponding to the occurrences of dissonant
intervals and unisons.

@@

The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.
As Figure~\ref{fig:motioncadence} left shows, there are four kinds of motion:
\emph{contrary} (two lines go in different directions),
\emph{parallel} (two lines go in the same direction by the same
distance), \emph{similar} (two lines go in the same direction by
different distances), and \emph{oblique} (one line plays the same note).
It is known that approaching a perfect interval by parallel or similar
motion tends to fuse the melodic lines into one.
To rule out such motion, we define the \texttt{checkMotion}
function, which inspects all pairs of two adjacent intervals (using an
auxiliary function \texttt{pairs}) and returns a list of erroneous patterns.

@@

The ending of the music should express relaxation.
There is a consensus that the unison and 8th are the most stable
intervals, hence the music must end with either of these intervals.
The last interval should also be approached by a \emph{cadence},
a progression that gives rise to a sense of resolution
(Figure~\ref{fig:motioncadence} right).
This in turn suggests that a valid ending requires the middle bars
to be non-empty.
We encode these rules as the \texttt{checkEnding} function, which,
upon finding an invalid ending, reports one of the three possible
errors.

@@

With this record type, we can show that the counterpoint in
Figure~\ref{fig:fs} is correct, since the music is an inhabitant of
\texttt{FirstSpecies}.

@@

In second species, the first and last bars may have a different structure
from middle bars.
More specifically, it is encouraged to begin the counterpoint line with
a half rest and end with a whole note.
Therefore, in our implementation, we reuse the \texttt{PitchInterval} 
type for the first and last bars, and define a new type
\texttt{PitchInterval2} for middle bars.

@@

middle2 : List PitchInterval2 -- Pitch \(\times\) Interval \(\times\) Interval
middle2 =
  (d 5 , min3 , per5) :: (e 5 , min3 , min6) :: (f 5 , maj3 , aug4) ::
  (e 5 , min6 , min3) :: (d 5 , min3 , maj6) :: []

@@

Strong beats in middle bars are constrained by the same rules as in
first species: they must all be consonant, non-unison intervals.
% We encode this restriction as the \texttt{checkStrongBeats}
% function, which, as the \texttt{checkIntervals} function for first
% species, returns a list of found errors.

@@

Parallel and similar motion towards a perfect interval is prohibited
across bars.
% We enforce this rule using the \texttt{checkMotion2} function,
% which uses \texttt{expandPitchIntervals2} to convert a list of
% \texttt{PitchInterval2} into a list of \texttt{PitchInterval}, and
% checks every pair of weak beat and strong beat intervals.

@@

The type-theoretical formalization of counterpoint has first been
attempted by \citet{szamozvancev-welltyped}.
They encode the rule for intervals by declaring a type class
\texttt{ValidHarmInterval} that classifies valid intervals, and then
specifying which instances are invalid using GHC's support for
custom type errors.
They also implement the restriction on motion based on a similar idea.

The availability of the custom type error feature is an advantage of
using Haskell: it allows the rule writer to leverage Haskell's type checker.
We believe that extending Agda with such support would be
beneficial in music formalization and for many other purposes.

@@

Our representation of counterpoint improves on Cong and Leo's
in two ways.
First, while Cong and Leo rely on the Agda type checker to report
errors, we use our own type checker (implemented as the
\texttt{checkXXX} functions) and type errors (defined as the
\texttt{XXXError} datatypes).
This allows us to produce more user-friendly error messages.
Suppose we have replaced the second interval of Figure~\ref{fig:fs} by
an octave (8th) of D, breaking the rule that perfect intervals cannot
be approached by parallel motion.
In our implementation, this causes an error in the \texttt{motionOk}
field, with the following message:

@@

\noindent In contrast, Cong and Leo would just report the existence
of an unsolved metavariable \texttt{\_13}, meaning that Agda failed
to construct a proof required by the counterpoint constructor.
Thus, the user would see the following message in the Agda buffer:

@@

The second improvement from \citet{cong-cp} is that, instead of
incorporating all the rules into a counterpoint datatype, we define
each rule as a separate function and combining them using a record type.
This makes it easy to switch between strict and relaxed rule sets:
we just need to add or remove certain fields.
It also allows us to reuse some of the rules in a context other than
counterpoint, as we will see in the next section.

>>>>

In this section, we present an implementation of species counterpoint,
based on the formulation given by \citet{fux-cp}.
The idea is to represent ``good'' counterpoint  as a dependent record,
whose fields encode the musical content as well as proofs that the
counterpoint follows certain rules.
Due to the space limitation, we only show some essential parts of
our implementation; see our anonymized supplement
\texttt{Counterpoint.agda} for full details.

@@

\begin{alltt}  
first : PitchInterval -- defined as Pitch \(\times\) Interval
first = (c 5 , per8)

@@

The middle bars of the music should maintain consonance of intervals
and independence of melodic lines.
As we saw in Figure~\ref{fig:interval}, consonant intervals include
the unison, 3rd, 5th, 6th, and 8th.
Among these, the unison is clearly an obstacle to distinguishing
between the two lines of music.
Therefore, the middle bars must consist only of the latter four intervals.
We encode this rule as the \texttt{checkIntervals} function, which
returns a list of errors corresponding to the occurrences of dissonant
intervals and unisons.

@@

The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.
As Figure~\ref{fig:motioncadence} left shows, there are four kinds of motion:
\emph{contrary} (two lines go in different directions),
\emph{parallel} (two lines go in the same direction by the same
distance), \emph{similar} (two lines go in the same direction by
different distances), and \emph{oblique} (one line plays the same note).
It is known that approaching a perfect interval by parallel or similar
motion tends to fuse the melodic lines into one.
To rule out such motion, we define the \texttt{checkMotion}
function, which inspects all pairs of two adjacent intervals (using 
auxiliary functions \texttt{motion} and \texttt{pairs}) and returns a
list of erroneous patterns.

@@

The ending of the music should express relaxation.
Since the unison and 8th are considered most stable, the music
must end with either of these intervals.
The last interval should also be approached by a \emph{cadence},
a progression that gives rise to a sense of resolution
(Figure~\ref{fig:motioncadence} right).
This in turn suggests that a valid ending requires the middle bars
to be non-empty.
We encode these rules as the \texttt{checkEnding} function, which,
upon finding an invalid ending, reports one of the three possible
errors.

@@

With this record type, we can show that the counterpoint in
Figure~\ref{fig:fs} is correct, by constructing the following proof term:

@@

In second species, the first and last bars may have a different structure
from the middle bars.
More specifically, it is encouraged to begin the counterpoint line with
a half rest and end with a whole note.
Therefore, in our implementation, we reuse the \texttt{PitchInterval} 
type for the first and last bars, and define a new type
\texttt{PitchInterval2} for the middle bars.

@@

middle2 : List PitchInterval2 -- defined as Pitch \(\times\) Interval \(\times\) Interval
middle2 =
  (d 5 , min3 , per5) :: (e 5 , min3 , min6) :: (f 5 , maj3 , aug4) ::
  (e 5 , min6 , min3) :: (d 5 , min3 , maj6) :: []

@@

Strong beats in the middle bars are constrained by the same rules
as in first species: they must all be consonant, non-unison intervals.
% We encode this restriction as the \texttt{checkStrongBeats}
% function, which, as the \texttt{checkIntervals} function for first
% species, returns a list of found errors.

@@

Parallel and similar motion towards a perfect interval is prohibited
across bars, that is, between the weak beat in one bar and the
strong beat in the next bar.
% We enforce this rule using the \texttt{checkMotion2} function,
% which uses \texttt{expandPitchIntervals2} to convert a list of
% \texttt{PitchInterval2} into a list of \texttt{PitchInterval}, and
% checks every pair of weak beat and strong beat intervals.

@@

The type-theoretical formalization of counterpoint has first been
attempted by \citet{szamozvancev-welltyped}.
They encode the rule for intervals by declaring a type class
\texttt{ValidHarmInterval} that classifies valid intervals, and then
specifying which instances are invalid using GHC's support for
custom type errors.
They also rule out invalid motion using a similar technique.

The idea of producing custom musical type errors underlies our
implementation as well.
The difference is that we have no means to leverage the Agda
type checker due to the lack of such a language feature, which
we consider to be a downside of using Agda for musical purposes.

@@

Our representation of counterpoint improves on Cong and Leo's
in two ways.
First, while they rely on the Agda type checker to report errors,
we use our own type checker (implemented as the
\texttt{checkXXX} functions) and type errors (defined as the
\texttt{XXXError} datatypes).
This allows us to produce more user-friendly error messages.
To see the difference, suppose we have replaced the second interval
of Figure~\ref{fig:fs} by an octave (8th) of D, breaking the rule
that perfect intervals cannot be approached by parallel motion.
In our implementation, this causes an error in the \texttt{motionOk}
field, with the following message:

@@

\noindent In contrast, Cong and Leo would just report the existence
of an unsolved metavariable, meaning that Agda failed to construct
a proof required by the counterpoint constructor.
Thus, the user would see the following message in the Agda buffer:

@@

The second improvement from \citet{cong-cp} is that, instead of
hard-coding all the rules into a datatype designed specifically for
counterpoint, we define each rule as a separate function and
combine them using a record type.
This makes it easy to switch between strict and relaxed rule sets:
we just need to add or remove certain fields.
It also allows us to reuse some of the rules in a context other than
counterpoint, as we will see in Section~\ref{sec:harmony:complete}.

### 2020-03-04 doc/icfp20/harmony.tex
final tweaks & swap figures

\newcommand{\Harmonization}{
  \begin{figure}
    \includegraphics[width=5.5cm]{fig/piston.png}
    \hspace{1cm}
    \includegraphics[width=5cm]{fig/fharm.png} \\
    \begin{flushleft}
    \begin{small}
     \hspace{2.35cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \ \ \,IV \,VI \ \,III \ IV \ \ \,V
     \hspace{2.6cm} V \,III \ \ I \ III \ \,II IV \,III\,IV \ \,V
    \end{small}
    \end{flushleft}
    \caption{Harmonization by Piston (left) and \fharm (right)}
    \Description{Harmonization by Piston (left) and \fharm (right)}
    \label{fig:harmonization}
  \end{figure}
}

\newcommand{\HarmonizationMT}{
  \begin{figure}
    \includegraphics[width=5.5cm]{fig/harmmt.png}
    \begin{flushleft}
    \begin{small}
     \hspace{5.4cm} I \ \ V \ \ \ VI \ \,V \ \ IV \ I \ \ \ V  \,IV \ \ \ \ V
    \end{small}
    \end{flushleft}
    \caption{Harmonization by Music Tools}
    \Description{Harmonization by Music Tools}
    \label{fig:harmonizationMT}
  \end{figure}
}

@@

Harmony refers to simultaneously sounding tones (chords) and its study
is primarily concerned with the progression of chords throughout a
piece of music. It is in essence the of dual of counterpoint. In this
section we formalize a small part of the classic text
\citet{piston-harmony}, in particular part of Chapter~9 which is
concerned with techniques for harmonizing a given melody.

@@

In Haskell this would be easy---there is an \texttt{Eq} type class and a
method for automatically generating equalities. In the Agda Standard
Library~\citep{agda-stdlib} list membership is a relation with
constructive evidence of membership, and requires far more effort than we
would like given that we currently don't need to do
proofs. Alternatives include defining decidable equality for triads
(which requires writing out 49 cases) or mapping to natural numbers
and using already implemented decidable or boolean equality for
those. For now we simply write the inverse function manually,
excluding the rare cases.

@@

As Piston does we require the harmonizing triad to include the melody
note. We use a function \texttt{harmonizations} to generate a list of
all candidate harmonizations for a given melody. For space
considerations we refer the reader to the supplement
\texttt{Harmony.agda} for its definition. Here again we encounter difficulties
with equality and membership. We use a bit vector to represent the
set of scale degrees in a triad, where each scale degree is mapped to
an element of \texttt{Fin 7}. However we are able to make use of
dependent types in the form of vectors to ensure the length of each
harmonization matches the melody, and continue to use vectors in the
following functions where appropriate.

@@

The harmonization produced is shown in Figure~\ref{fig:harmonizationMT}.

\HarmonizationMT

@@

Most of the Haskell-based
work~\citep{magalhaes-harmtrace,koops-fharm,magalhaes-fcomp} has
focused on harmony. In particular \fharm~\citep{koops-fharm} tackles
the problem of harmonization of a melody;
Figure~\ref{fig:harmonization} right shows their harmonization of
Piston's example. Their method is quite different from
ours. Instead of using Piston's table of root progressions they use a
more sophisticated hierarchical grammar from previous
work~\citep{magalhaes-harmtrace}.
More specifically, they first require the melody notes to be members of
the harmonizing chords, then use their previous work to try to create
a harmonic analysis of every possible sequence of chords that arises,
choosing the one that most closely fits their grammar.

Although the result produced by \fharm is not bad, it clearly leaves
room for improvement. Noted in
the paper is that no attention is given to voicing (how each of the
four parts flows horizontally and how the lines interact---in other
words the counterpoint). They use inversions (in which the
third or fifth of the chord is in the bass) to ensure each melody line
does not move too much, but as the harmonizing triads are all in close
position separated from the melody, the result sounds like a soprano
melody with block chord accompaniment, rather than four independent
lines.

Furthermore, \fharm always double the melody note, not the root note
as Piston does (and as preferred in four-part harmony in general).
Also, minimizing the number of parse errors as a means
to choose the best harmonization does not necessarily have a musical
meaning.

>>>>

\newcommand{\Harmonization}{
  \begin{figure}
    \includegraphics[width=5.5cm]{fig/piston.png}
    \hspace{1cm}
    \includegraphics[width=5.5cm]{fig/harmmt.png} \\
    \begin{flushleft}
    \begin{small}
     \hspace{2.1cm} V \ \ \ \ \ \ \ \,I \ \ \ \ \ \ \ \,IV \,VI \ \,III \ IV \ \ \,V
     \hspace{2.65cm} I \ \ V \ \ \ VI \ V \ \ IV \ I \ \ \ V  \,IV \ \ \ \ V
    \end{small}
    \end{flushleft}
    \caption{Harmonization by Piston (left) and Our Implementation (right)}
    \Description{Harmonization by Piston (left) and Our Implementation (right)}
    \label{fig:harmonization}
  \end{figure}
}

\newcommand{\HarmonizationF}{
  \begin{figure}
    \includegraphics[width=5.5cm]{fig/fharm.png}
    \begin{flushleft}
    \begin{small}
     \hspace{5.5cm} V \,III \ \ \ I \ \,III \ \,II \,IV \ \,III IV \ \ V
    \end{small}
    \end{flushleft}
    \caption{Harmonization by \fharm}
    \Description{Harmonization by \fharm}
    \label{fig:harmonizationF}
  \end{figure}
}

@@

Harmony refers to simultaneously sounding tones (chords) and its study
is primarily concerned with the progression of chords throughout a
piece of music. It is in essence the dual of counterpoint. In this
section we formalize a small part of the classic text
\citet{piston-harmony}, in particular part of Chapter~9 which is
concerned with techniques for harmonizing a given melody.
The complete implementation can be found in \texttt{Harmony.agda}
and \texttt{Piston.agda} of the anonymized supplement.

@@

In Haskell this would be easy---there is an \texttt{Eq} type class and a
method for automatically generating equalities. In the Agda Standard
Library~\citep{agda-stdlib} list membership is a relation with
constructive evidence of membership, and requires far more effort than we
would like, given that we currently do not need to do
proofs. Alternatives include defining decidable equality for triads
(which requires writing out 49 cases), or mapping to natural numbers
and using already implemented decidable or boolean equality for
those. For now we simply write the inverse function manually,
excluding the rare cases.

@@

As Piston does we require the harmonizing triad to include the melody
note. We use a function \texttt{harmonizations} to generate a list of
all candidate harmonizations for a given melody (see
\texttt{Harmony.agda} for its definition). Here again we encounter difficulties
with equality and membership. We use a bit vector to represent the
set of scale degrees in a triad, where each scale degree is mapped to
an element of \texttt{Fin 7}. However we are able to make use of
dependent types in the form of vectors to ensure the length of each
harmonization matches the melody, and continue to use vectors in the
following functions where appropriate.

@@

The harmonization produced is shown in Figure~\ref{fig:harmonization}
right.

@@

\HarmonizationF

Most of the Haskell-based work on
music~\citep{magalhaes-harmtrace,koops-fharm,magalhaes-fcomp} has
focused on harmony. In particular \fharm~\citep{koops-fharm} tackles
the problem of harmonizing a melody;
Figure~\ref{fig:harmonizationF} shows their harmonization of
Piston's example. Their method is quite different from
ours. Instead of using Piston's table of root progressions they use a
more sophisticated hierarchical grammar from previous
work~\citep{magalhaes-harmtrace}.
More specifically, they first require the melody notes to be members of
the harmonizing chords, then use their previous work to try to create
a harmonic analysis of every possible sequence of chords that arises,
choosing the one that most closely fits their grammar.

Although the result produced by \fharm is not bad, it clearly leaves
room for improvement. Noted in
the paper is that no attention is given to voicing (how each of the
four parts flows horizontally and how the lines interact---in other
words the counterpoint). They use inversions to ensure each melody line
does not move too much, but as the harmonizing triads are all in close
position separated from the melody, the result sounds like a soprano
melody with block chord accompaniment, rather than four independent
lines.

Among those issues not discussed in the paper, one is the choice of
notes to be doubled: \fharm always doubles the melody note, while
in four part harmony it is generally preferred to double the root
note as Piston does.
Another concern is that minimizing the number of parse errors as a means
to choose the best harmonization does not necessarily have a musical
meaning.

### 2020-03-04 doc/icfp20/music.tex
final tweaks & swap figures

For rhythm, the fundamental concept is \emph{duration}, an
unspecified unit of time during which a sound or silence lasts.
We represent duration as a natural number, and use this value to
calculate the absolute length when the music is played at a specific
tempo.
As an example, the duration constant \texttt{whole} has value
\texttt{16} and corresponds to 4 seconds if the tempo is 60 beats
per minute.

@@

An \emph{interval} represents the difference in pitch between
two notes.
There are 13 kinds of interval within an octave
(Figure~\ref{fig:interval}), and these intervals can be classified from
several different perspectives:
(i) major or minor; (ii) consonant or dissonant; and (iii) perfect
or imperfect.
We represent intervals as natural numbers, but in this paper, we
refer to them using names such as \texttt{maj3} and \texttt{per8},
which are more reader-friendly.

@@

A \emph{chord} refers to a set of simultaneously sounding notes.
In this paper we are particularly interested in \emph{triads}, which are
chords consisting of three notes separated by the 3rd intervals (i.e.,
\texttt{maj3} and \texttt{min3}).
We define triads as a datatype with roman numeral constructors
representing the scale degree of the lowest note (called the \emph{root}).
The other two notes are then members of the scale a third and a fifth
above the root.
For instance, the triad \texttt{I} in C major corresponds to the chord
C-E-G.
Triads can also be arranged in a way that a non-root note appears as
the lowest note.
Such arrangement is called \emph{inversion}; if a triad is not applied
inversion, it is in \emph{root position}.

>>>>

For rhythm, the fundamental concept is \emph{duration}, an
unspecified unit of time during which a sound or silence lasts.
Examples of duration include whole, half, and quarter;
we define them simply as natural number constants.

@@

An \emph{interval} represents the difference in pitch between
two notes.
There are 13 kinds of interval within an octave
(Figure~\ref{fig:interval}), and these intervals can be classified from
several different perspectives:
(i) major or minor; (ii) consonant or dissonant; and (iii) perfect
or imperfect.
We represent intervals as natural numbers, and refer to them using
conventional names such as \texttt{maj3} and \texttt{per8}.

@@

A \emph{chord} refers to a set of simultaneously sounding notes.
Of our particular interest is \emph{triads}, which are
chords consisting of three notes separated by the 3rd intervals (i.e.,
\texttt{maj3} and \texttt{min3}).
We define triads as a datatype with roman numeral constructors
representing the scale degree of the lowest note (called the \emph{root}).
The other two notes are then members of the scale a third and a fifth
above the root.
For instance, the triad \texttt{I} in C major corresponds to the chord
C-E-G.
Triads can also be arranged in a way that a non-root note appears as
the lowest note.
Such arrangement is called \emph{inversion}; if a triad is not applied
inversion, we say that it is in \emph{root position}.
